---
author:
  - name: Derek Corcoran
    affiliation: University of Missouri
    address: >
      First line
      Second line
    email: corcoranbarriosd@missouri.edu
    url: http://rstudio.com
  - name: Elisabeth Webb
    affiliation: University of Missouri, USGS
  - name: Dylan Kesler
    affiliation: Institute for Bird Populations
title:
  formatted: "\\pkg{DiversityOccupancy}: A package to model occupancy, abundance, diversity and selection of prioriry areas based on repeated presence-absence sampling"
  short:     "\\pkg{DiversityOccupancy}: Selecting Priority Areas"
abstract: >
  Occupancy modeling has been widely used as a tool for ecological research and management planning in the context of single species models. We present the package \pkg{DiversityOccupancy} in the \proglang{R} environment. The objective of this package is to simultaneously model factors associated with occupancy and abundance of individual species using a detection history to model occupancy and/or abundance. This new package also uses predicted abundances to calculate species diversity for each sampling site. \pkg{DiversityOccupancy} models factor(s) associated with among-site species diversity, which can be combined with spatial data to identify areas that contain both high abundance of species of conservation concern and high species diversity.
  
keywords:
  # at least one keyword must be supplied
  formatted: [ "\\pkg{DiversityOccupancy}", Occupancy Modeling, "\\proglang{R}", alpha diversity, Protected areas]
  plain:     [keywords, not capitalized, Occupancy modeling]
preamble: >
  \usepackage{amsmath}
  \usepackage{amsfonts}
biblio-style: jss
output:
  rticles::jss_article:
    fig_caption: true
    citation_package: natbib
bibliography: Derek.bib
---


```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(rticles)
library(rmarkdown)
library(yaml)
library(xtable)
```

```{r, echo=FALSE, warning=FALSE}
library(knitr)
opts_chunk$set(prompt=TRUE)
options(replace.assign=TRUE, width=90, prompt="R> ")
knitr::opts_chunk$set(warning =  FALSE)
```

# Introduction

##Imperfect detection in wildlife surveys

One of the main objectives of ecology is to study the abundance and distribution of species and populations. The problem is that estimation of population is time consuming, hard to replicate and almost impossible in rare species due to the fact that they are hard to detect, which leads to false absences that might cause underestimation of abundances and distribution. Despite this, most sampling and analysis techniques don’t take into account the inherent imperfect detection found in natural populations. One of the tools used to overcome this limitation is occupancy modeling.

##Occupancy models, taking into account detection probability

In the last decade, Occupancy modeling has been increasingly used as a method to account for how species respond to environmental or anthropogenic factors. It has also been shown to be useful as a species distribution-modeling tool when species have imperfect detection. Another use for what it has been used is for managers to change the environment of managed areas in order to improve the status of species of conservation concern \citep{mackenzie_estimating_2002, mackenzie2006occupancy} but as far as we know this is the first method that takes into account both species diversity and individual species abundance in order to select conservation areas.
Occupancy models use detection-nondetection data from repeat surveys to simultaneously estimate probabilities of detection (p) and occupancy $\psi$ (Kellner and Swihart 2014, MacKenzie 2006). Occupancy is defined as the proportion of habitat occupied by a species (MacKenzie et al. 2002)


##Occupancy modeling y formulacion matematica


## Single-species or multiple-species management

The ecological and management literature has historically valued the idea of managing for diversity. This, however, clashes with the classic way in which conservation takes place. Conservation agencies, governments, scientists and international organizations such as International Union for Conservation of Nature (IUCN), classify species according to a conservation status, and policies are enacted to safeguard the ones that species of conservation concern \citep{keller2004red, rodrigues2006value}. Managing for single species is often less complicated than simultaneously addressing suites of species, and it is may be simpler to track the status of a species than it is to define management for diversity \citep{simberloff1998flagships}. Further, tracking the conservation status of a single species is often more straightfroward than tracking multiple species or diversity.

There are several approaches to designing management measures based on single species, such as using umbrella species \citep{crosby2015looking, bichet2016maintaining}, Habitat Suitability Indices (HSI) \citep{reza2013integrating, soniat2013predicting, zohmann2013modelling} and Species distribution modeling (SDM) \citep{peterson2011ecological, guisan2013predicting} among others. Even when the literature mostly criticizes the use of such approaches due to the ineffectiveness in preventing loss of biodiversity \citep{roberge2004usefulness, branton2011assessing}.

One of the problems of managing simple species, is that is usually very difficult to predict what will happen to other species and because of that it is hard to predict what will happen to diversity \citep{pulliam2000relationship}, with some examples showing that even management measurements that use some species as umbrella species for conservation of ecosystems leading to undesired community effects \citep{white2013conservation}.

Managing for multiple-species, although desirable, it has shown to be difficult to implement \citep{mollmann2014implementing, lmgren2015baltic}, most multimodel species only take into account only the number of species in an area, not accounting for how rare or common some species are, or they do not take into account the presence of some endangered species, since they are only trying to account for higher number of species, and not their identity \citep{taft2002waterbird, tori2002wetland, plaganyi2014multispecies}

There are scenarios where a manager could want to manage for biodiversity, but in many countries, laws such as the Endangered Species Act in the United States or Canada's Species at Risk Act \citep{congress1973endangered, waples2013tale}, will require the manager or scientist to focus at the species level.The package presented in this article pretends to be a tool to change this either/or scenario and take information of both diversity and individual species models.

In this paper we present package \pkg{DiversityOccupancy}, used in the \proglang{R} environment, The objective of this package is to simultaneously model factors associated with occupancy and abundance of individual species using a detection history file, and to use predicted abundances to calculate species diversity for each sampling site. The package then models factor(s) associated with among-site species diversity, which can then be combined with spatial data to identify areas that contain both high abundance of species of conservation concern and high species diversity.


In the last decade, Occupancy modeling has been used more and more as a method to account for how species respond to environmental or anthropogenic factors. It has also been shown to be useful as a species distribution modeling tool when species have imperfect detection. Another use for what it has been used is for managers to change the environment of managed areas in order to improve the status of species of conservation concern \citep{mackenzie_estimating_2002, mackenzie2006occupancy} but as far as we know this is the first method that takes into account both species diversity and individual species abundance in order to select conservation areas.

Occupancy models use detection-nondetection data from repeat surveys to simultaneously estimate probabilities of detection (p) and occupancy ($\psi$) \citep{mackenzie2006occupancy}. 
\citep{burnham2003model}.  We standardized all continuous site covariates t


```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)

```

## Occupancy modeling

### Imperfect detection and the need for occupancy modeling

One of the main objectives of ecology is to study the abundance and distribution of species and populations. The problem is that estimation of population is time consuming, hard to replicate and almost impossible in rare species due to the fact that they are really hard to detect, whic leads to false absences that might bias estimates. Despite this most sampling and analysis techniques don't deal with imperfect detection. Occupancy models use detection-nondetection data from repeat surveys to simultaneously estimate probabilities of detection and occupancy \citep{kellner2014accounting}. Occupancy is defined as the proportion of habitat occupied by a species \citep{mackenzie_estimating_2002}

### History of occupancy modeling

Occupancy modeling was first proposed in 2002 \citep{mackenzie_estimating_2002}

### Math of Occupancy modeling

Lets suppose we are trying to determine the probability of a species presence in a site ($\psi$), given the probability of detection (p). There are t Trials in the field to detect the species in a site, and the detections (1) and non-detections (0) are tallied as a detection history. Then after the estimation of p, we can say that the cumulative probability of detection is:

$$
\begin{aligned}
  \ p* = 1 -  \left( 1 - p \right)^t \
\end{aligned}
$$

If S sites are sampled and the species was detected Sd times, our estimation of $\psi$ is:

$$
\begin{aligned}
  \psi & = \frac{Sd}{S \times\ p*} \
\end{aligned}
$$


Furthermore, both the probability of detection (p) and occupancy ($\psi$) may be a function of covariates. We can extent the model to include covaiates using the following formula.

$$
\begin{aligned}
  \psi = \frac{exp(Cov \times\ \beta)}{1 + exp(Cov \times\ \beta)} \
\end{aligned}
$$

### History of occupancy modeling software's and it's limitation (the need for DiversityOccupancy)

### Comparison of DiversityOccupancy and other occupancy softwares (TABLE)

- camptrapR (\proglang{R})
- downscale (\proglang{R})
- hillmakerR (\proglang{R})
- pom (\proglang{R})
- Presence
- stocc (\proglang{R})
- Unmarked (\proglang{R})

```{r, echo=FALSE, results='asis'}
library(xtable)
Softwares <- structure(list(X = structure(c(9L, 5L, 7L, 4L, 1L, 8L, 6L, 2L, 
3L), .Label = c("Alpha diversity models", "Build in model selection", 
"Built in response plot", "Multiple Species Abundance", "Multiple Species Single Season occupancy", 
"Priority area selection", "Single Species Abundance", "Single Species Dynamic occupancy", 
"Single Species Single Season occupancy"), class = "factor"), 
    Diversityoccupancy = structure(c(1L, 1L, 1L, 1L, 1L, NA, 
    1L, 1L, 1L), .Label = "$\\checkmark$", class = "factor"), Unmarked = structure(c(1L, 
    NA, 2L, NA, NA, 2L, NA, NA, NA), .Label = c("$\\checkmark$", 
    "$\\checkmark$"), class = "factor"), stocc = structure(c(1L, NA, NA, 
    NA, NA, NA, NA, NA, NA), .Label = "$\\checkmark$", class = "factor"), 
    Presence = structure(c(1L, NA, 1L, NA, NA, 1L, NA, NA, NA
    ), .Label = "$\\checkmark$", class = "factor"), Pom = structure(c(1L, 
    NA, NA, NA, NA, NA, NA, NA, NA), .Label = "$\\checkmark$", class = "factor"), 
    camptrapR = structure(c(1L, NA, NA, NA, NA, NA, NA, NA, NA
    ), .Label = "$\\checkmark$", class = "factor")), .Names = c("X", "Diversityoccupancy", 
"Unmarked", "stocc", "Presence", "Pom", "camptrapR"), class = "data.frame", row.names = c(NA, 
-9L))

names <- colnames(Softwares[,-1])
colnames(Softwares) <- c("", names)
print(xtable(Softwares, align = rep("c", 8)), scalebox='0.75', type = "latex", sanitize.text.function = function (x) x, include.rownames = F, comment=FALSE)


```


## Installing DiversityOccupancy

### Requirements

To use this package you need \proglang{R} version 3.2.2 or newer (use the function \code{sessionInfo()} in your R session to check your current version).

### Installing the package

Install from cran repository

```{r, echo=TRUE, eval=FALSE}
install.packages("DiversityOccupancy")
```

### Objectives of the Package

### Multiple species occupancy modeling

### Abundance from multiple species occupancy modeling

### Diversity from abundance

### Spacially explicit predictions

### Selection of priority areas

### Graphical advantages from the package

### Example dataset

As an example we probide part of the  dataset collected in Pohnpei Island \citep{oleiro2014avian}, containing the detection history of five species in four consecutive days in 120 different locations of the Island. 

## Pohnpei in the federal states of micronesia

The Island of Pohnpei is a 334 squared kilometer island, in Oceania. it is the largest and most populated Island in the Federal States of Micronesia. Like every oceanic Island it has high endemism. It has a rich mosaic of ecosystems composed by manglars, palm forest, dwarf forests among others \citep{raynor1994resource, buden2000comparison, merlin2005kava}.

## Bird Species

# Use of the package

In order to calculate abundance and alpha diversity we need at least three files:

#### Detection history of multiple species

A data frame consisting on the detection history of at least two species.  As an example \pkg{DiversityOccupancy} has the data-set Islandbirds which contains detections history of 5 species in the Pohnpei Island for 4 consecutive days (Columns) in 120 different sites (Rows). The data set includes a 1 for each time a species was detected, and a 0 for each time it was not detected.

A detection for the first two species is presented below:

```{r, echo=TRUE, message=FALSE}
library(DiversityOccupancy)
data("IslandBirds")
head(IslandBirds[1:8], 10)
```

####Site covariates

Site covariates are presented in a data frame consisting of measurements taken at each site. The covariates are used singly and in combination to model occupancy or abundance, and they should be variables that are stable within the scope of the length of the study. In \pkg{DiversityOccupancy} there is an example concordant with the IslandBirds data set called siteCov:

```{r, echo=TRUE, eval=FALSE}
data("siteCov")
head(siteCov, 10)
```


```{r, echo=FALSE}
data("siteCov")
head(siteCov)
```

####Detection covariates

A list of data frames, in which each data frame includes a daily measurement of variables with the potential to affect detection probabilities. It is important that each element (data frame) of the list has a name, so that it can be called to fit the occupancy model. These variables are used to model the probability of detection.

\pkg{DiversityOccupancy} has a data set called *Daily_Cov* which illustrates how the Daily covariates have to be structured:

```{r, echo=TRUE}
#All the items of the ist must have names
names(Daily_Cov)
#here we see the first dataframe of the Daily_Cov dataset
head(Daily_Cov[[1]])
```

##Fiting models for occupancy

As an example of batch fitting occupancy models, we will use the *IslandBirds* dataset, at first we will do a simple occupancy model without automatic model selection:

```{r, echo=TRUE, cache=TRUE, message=FALSE, warning=FALSE}
BirdOccupancy <-batchoccu(pres = IslandBirds, sitecov = siteCov, obscov = Daily_Cov, spp =  5, form = ~ Day + Wind + Rain + Noise + Clouds ~
Elev + AgroFo + SecVec + Wetland)
```

In this case *pres* expects the detection history file explained above, *sitecov*, expects a dataframe with the covariates used to estimate occupancy, *obscov* is a list of dataframes with the variables expected to affect the probability of detection, *spp* is the number of species in the *pres* file, and *form* is the formula used to fit the model  

This creates an object of the class *batchoccupancy* We can check the fitted model accesing the *models* element

```{r, echo=TRUE, cache=TRUE, message=FALSE, warning=FALSE}
BirdOccupancy$models
```

In order to do automatic model selection for the occupancy models, you have to change the option of dredge to TRUE:

```{r, echo=TRUE, cache=TRUE, message=FALSE, warning=FALSE}
BirdOccupancy <-batchoccu(pres = IslandBirds, sitecov = siteCov, obscov = Daily_Cov, spp =  5, form = ~ Day + Wind + Rain + Noise + Clouds ~
Elev + AgroFo + SecVec + Wetland, dredge = TRUE)
```

###Ploting the response of occupancy for each species and each variable

\pkg{DiversityOccupancy}, has a build in function *responseplot.occu* that will automatically plot the occupancy responce to the variation of any of the covariates from the *sitecov* file. To use this function you need a *batchoccupancy* model to feed the batch parameter, and then you have to select which species and variable to plot. In the next graph, we are ploting the responce of the fourth species (*Rukia longirostra*), for elevation. This function automatically limits the variable of interest between the minimum and maximum value found in the field.

```{r, echo=TRUE, cache=TRUE, message=FALSE, warning=FALSE}
responseplot.occu(batch = BirdOccupancy, spp = 4, variable = Elev)
```

This next plot shows again, the response to elevation, but in this case for species 5 (*Ducula oceanica*).

```{r, echo=TRUE, cache=TRUE, message=FALSE, warning=FALSE}

responseplot.occu(batch = BirdOccupancy, spp =  5, variable = Elev)
```

##Fiting models for abundance and predicting alpha diversity

In this example we will fit and model the abundance for 5 bird species and calculate alpha diversity from those results.

```{r, echo=TRUE, cache=TRUE, message=FALSE, warning=FALSE}
birdDiversity <-diversityoccu(pres = IslandBirds, sitecov = siteCov,
obscov = Daily_Cov,spp =  5, form = ~ Day + Wind + Time ~ Elev + Wetland + Upland, dredge = FALSE)
```

The resulting object of class diversityoccupancy has the following elements

```{r, echo=TRUE, message=FALSE, warning=FALSE}
names(birdDiversity)
```

If you need to see the parameters of the model of one of the species, you call the species number with the element$models. For example extract the model for the second species:

```{r, echo=TRUE, message=FALSE, warning=FALSE}
birdDiversity$models[[2]]
```

The species parameter for a diversityoccupancy object shows us a table with the abundance and alpha diversity calculated for each sampled point:

```{r, echo=TRUE, message=FALSE, warning=FALSE, eval=TRUE}
summary(birdDiversity$species)
```

## Automatic model selection for abundance models

If the option of dredge is set to “TRUE”, then diversityoccu attempts to fit all first order models, and it selects the one with the lowest AICc value, for each species.  Be aware that processing times rapidly increases with added numbers of parameters, and that processing can require many hours or days for complex data sets.  The following graph and table shows the processing time for the IslandBirds data set.

```{r, echo=FALSE, cache=TRUE, warning=FALSE, message=FALSE}
time.comp<- read.csv("~/DiversityOccupancyJSS/DiversityOccuPaper/Times.csv")
library(ggplot2)
ggplot(time.comp, aes(x = K, y = Time)) + geom_smooth() + geom_point()+ggtitle("Time taken for model selection") + labs(x = "Number of Parameters [K]" , y = "Processing time [seconds]") + theme_bw()
```

From now on we will work with automatically selected models for bird abundance and diversity using an information theoretic approach (AICc).

```{r,echo=TRUE, cache=TRUE, warning=FALSE, message=FALSE, eval=TRUE}
birdmodel.selected <- diversityoccu(pres = IslandBirds, sitecov = siteCov,
obscov = Daily_Cov,spp =  5, form = ~ Day + Wind + Time ~ Elev + Wetland + Upland, dredge = TRUE)
```

Below we present an example of an analysis with the full model (includes all variables) and subsequently results from a model selection analysis, both of them only for the second species:

```{r, warning=FALSE, message=FALSE}
birdDiversity$models[[2]]
birdmodel.selected$models[[2]]
```

The responses of individual species to specific variables can be shown using the function responseplot.abund.abund, bellow we show the response of abundance in species 2 to the Burn intensity soil. Note that this function automatically bounds the limits of the variable to the maximum and minimum observable values in the field.

```{r, warning=FALSE, message=FALSE}
responseplot.abund(birdmodel.selected, spp = 2, variable = Elev)
```

##Model selection for alpha diversity modeling

The next step is to select the best model for predicting alpha diversity using the model.diversity function.  The function takes a diversityoccupancy object, and fits all possible glm models and ranks them by AICc. Other than the diversityoccupancy object, there are three other parameters to select: 1) Method, which can be either “h” which fits every possible model, or “g”, which uses genetic algorithms to select models (recommended for large candidate sets); 2) Delta, which allows the user to identify an AICc delta threshold, which returns all models with AICc values below the threshold; 3) Squared, which includes only linear combinations when set to FALSE (Default), and both linear and quadratic (second order) if set to TRUE.

```{r, warning=FALSE, message=FALSE, results='hide', cache=TRUE}
glm.BirdDiverse <- model.diversity(birdmodel.selected, method = "g", delta = 2, 
squared = TRUE)
```

To see the top models extract the Table element of the modelselection object

```{r, echo=FALSE}
glm.BirdDiverse$Table
```

The responseplot.diver function takes a modeldiversity object and one of the variables used to predict the alpha diversity index, and makes a plot showing the response of the diversity index against the selected variable. This function automatically limits the values of that variable to the maximum and minimum values of the dataset. It also shows the standard deviation of the estimation.

```{r, echo=TRUE}
responseplot.diver(glm.BirdDiverse, Elev)
```

Also since the returned plot is a ggplot type object, it can be easily modified following ggplot2 grammar of graphics.

```{r, echo=TRUE}
library(ggplot2)
k <- responseplot.diver(glm.BirdDiverse, Elev)
k
k + theme_gray()
k + ylim(c(0,1))
```

##Selecting conservation areas based on alpha diversity and abundance of species of conservation concern

Of the 5 species modeled, lets say that there are at least two of conservation concern, the second and thirds of our list.
Since we already have models relating site characteristics to species abundance and a model relating site characteristics to alpha diversity, with the use of a spatial raster layers (rasterstack) with the modeled variables we can choose an area with high species diversity and/or abundance. In order for this function to work properly the stack has to be in lon lat projection, and Birdstack is in UTM, so we reproject the data.

```{r, warning=FALSE}
library(raster)
data(Birdstack)
plot(Birdstack)
newproj <- '+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0'
library(rgdal)
Birdstack <- stack(projectRaster(Birdstack, crs=newproj))
```

In order to find areas of high conservation value, we use the predict.diversity function. We need both a diversityoccupancy and a modeldiversity class object, used in the model, and diverse parameters respectfully, a spatial representation of site covariates as raster files (rasterstack), with the variables in the new.data parameter, a boolean vector in the species parameter indicating which species shall be considered (T or F), and the quantile.nth parameter, which indicates a quantile threshold that is used for abundance and/or richness to indicate conservation value (areas above the threshold will be returned).



```{r, cache=TRUE, warning=FALSE, message=FALSE, results='hide', echo=TRUE}
Selected.area <- diversity.predict(model = birdmodel.selected, diverse = glm.BirdDiverse, new.data = Birdstack, quantile.nth = 0.65, species =
c(F,T,T,F,F))
```


From the object Selected.area we can extract and plot not only the area that was selected with the atributes, but also the layers that shoe the predicted diversity for the island and also the predicted abundance for each of the species that were used for the analysis.




#Discussion

# References